---
title: "The Influence of Text Characteristics for Email Classification"
author: "Group 22"
number-sections: true
format: 
  html:
    embed-resources: true
    code-tools: true
  pdf: default
editor_options: 
  chunk_output_type: console
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

```{r}
#| label: libraries
library(ggplot2)
#library(dplyr)
#library(moderndive)
#library(gapminder)
#library(skimr)
#library(tidyverse)
library(gt)
library(patchwork)
library(gridExtra)
library(broom)
library(knitr)
library(GGally)
library(sjPlot)
```

```{r}
#| label: read data
email<-read.csv("C:/Users/70652/Desktop/STATS5085 Data Analysis Skills/Project 2/DAS-Group-22/dataset22.csv")
```

```{r}
#| label: convert cht to factor
email$yesno<-as.factor(email$yesno)
```
## Introduction

This analysis focuses on identifying the characteristics of email messages that are predictive of spam. With the rise of unwanted and potentially harmful emails, effective spam detection has become a crucial challenge. By analysing patterns in email content, structure, and metadata, we aim to identify the distinguishing features that separate spam from legitimate messages. Gaining a deeper understanding of these characteristics can lead to the development of smarter, more accurate filtering systems that reduce inbox clutter and protect users from scams and other malicious activities. 





## Exploratory Data Analysis {#sec-EDA}

### Correlation

```{r}
#| label: fig-pairs
#| fig-cap: Correlations between each variables.
ggpairs(email[,1:6]) +
  theme(plot.background = element_rect(
    fill = "transparent",
    colour = NA,
    size = 1))
```

### Data Visualization

```{r}
#| label: fig-boxplot.crl.tot
#| fig-cap: Boxplot of total length of uninterrupted sequences of capitals.
ggplot(email, aes(x = yesno, y = crl.tot)) +
  geom_boxplot() +
  labs(x = "Spam indictor", y = "Uninterrupted sequences of capitals", 
       title = "Spam indictor with total length of uninterrupted sequences of capitals")
```

```{r}
#| label: fig-boxplot.dollar
#| fig-cap: Boxplot of occurrences of the dollar sign.
ggplot(email, aes(x = yesno, y = dollar)) +
  geom_boxplot() +
  labs(x = "Spam indictor", y = "Occurrences of the dollar sign", 
       title = "Spam indictor with occurrences of the dollar sign")
```

```{r}
#| label: fig-boxplot.bang
#| fig-cap: Boxplot of occurrences of '!'.
ggplot(email, aes(x = yesno, y = bang)) +
  geom_boxplot() +
  labs(x = "Spam indictor", y = 'Occurrences of "!"', 
       title = 'Spam indictor with occurrences of "!"')
```

```{r}
#| label: fig-boxplot.money
#| fig-cap: Boxplot of occurrences of "money".
ggplot(email, aes(x = yesno, y = money)) +
  geom_boxplot() +
  labs(x = "Spam indictor", y = 'Occurrences of "money"', 
       title = 'Spam indictor with occurrences of "money"')
```

```{r}
#| label: fig-boxplot.n000
#| fig-cap: Boxplot of occurrences of '000'.
ggplot(email, aes(x = yesno, y = n000)) +
  geom_boxplot() +
  labs(x = "Spam indictor", y = 'Occurrences of "000"', 
       title = 'Spam indictor with occurrences of "000"')
```

```{r}
#| label: fig-boxplot.make
#| fig-cap: Boxplot of occurrences of 'make'.
ggplot(email, aes(x = yesno, y = make)) +
  geom_boxplot() +
  labs(x = "Spam indictor", y = 'Occurrences of "make"', 
       title = 'Spam indictor with occurrences of "make"')
```

## Formal Data Analysis

### Fitting Model in Log-odds

```{r}
#| label: fiting model1
model1 <- glm(yesno ~ crl.tot+dollar+bang+money+n000+make, data = email, 
             family = binomial(link = "logit"))
```

```{r}
#| label: summary model1
summary(model1)
mod1coefs <- round(coef(model1), 2)
```

```{r}
#| label: fiting model2
model2 <- glm(yesno ~ crl.tot+dollar+bang+money+n000, data = email, 
             family = binomial(link = "logit"))
```

```{r}
#| label: summary model2
summary(model2)
mod2coefs <- round(coef(model2), 3)
```

```{r}
#| label: Find out the baseline category.
levels(email$yesno)
```



\begin{align}
\ln\left(\frac{p}{1-p}\right) &= \alpha + \beta_{crl.tot} \cdot \textrm{crl.tot} +\beta_{dollar} \cdot \textrm{dollar} +\beta_{bang} \cdot \textrm{bang} +\nonumber\\&\hspace{0.5cm}\beta_{money} \cdot \textrm{money} +\beta_{n000} \cdot \textrm{n000} \nonumber\\&= `r mod2coefs[1]` + `r mod2coefs[2]` \cdot \textrm{crl.tot} +`r mod2coefs[3]` \cdot \textrm{dollar} +`r mod2coefs[4]` \cdot \textrm{bang} +`r mod2coefs[5]` \cdot \textrm{money} +`r mod2coefs[6]`\cdot \textrm{n000} \nonumber
\end{align}








```{r}
#| label: summary confident of model coefficient
confint(model2) %>%
  kable()
```

```{r}
#| label: fig-CI-of-Log-odds
#| fig-cap: 95% CI of confidents in Log-odds.
plot_model(model2, show.values = TRUE, transform = NULL,
           title = "Log-Odds (Email indicator)", show.p = TRUE)
```



### Odds
We can obtain the odds scale $\frac{p}{1-p}$ by

$$
\frac{p}{1-p}= exp(`r mod2coefs[1]` + `r mod2coefs[2]` \cdot \textrm{crl.tot} +`r mod2coefs[3]` \cdot \textrm{dollar} +`r mod2coefs[4]` \cdot \textrm{bang} +`r mod2coefs[5]` \cdot \textrm{money} +`r mod2coefs[6]`\cdot \textrm{n000})
$$





```{r}
#| label: Getting odds scale.
model2 %>%
 coef() %>%
  exp()
```



```{r}
#| label: fig-CI-of-odds-scale
#| fig-cap: 95% CI of confidents in odds.
plot_model(model2, show.values = TRUE,
           title = "Odds (Email indicator)", show.p = FALSE)
```


### Probabilities
We can obtain the probability $p=\text{Prob}(\text{spam})$ by
$$
p=\frac{exp(`r mod2coefs[1]` + `r mod2coefs[2]` \cdot \textrm{crl.tot} +`r mod2coefs[3]` \cdot \textrm{dollar} +`r mod2coefs[4]` \cdot \textrm{bang} +`r mod2coefs[5]` \cdot \textrm{money} +`r mod2coefs[6]`\cdot \textrm{n000})}{1+exp(`r mod2coefs[1]` + `r mod2coefs[2]` \cdot \textrm{crl.tot} +`r mod2coefs[3]` \cdot \textrm{dollar} +`r mod2coefs[4]` \cdot \textrm{bang} +`r mod2coefs[5]` \cdot \textrm{money} +`r mod2coefs[6]`\cdot \textrm{n000})}
$$


## Conclusion
